Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands

### 2023-03-08 manuscript.Rmd
Update manuscript.Rmd

---
title: "Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands"
author: Markus Bauer*, Jakob Huber, Johannes Kollmann
output:
  officedown::rdocx_document:
    reference_docx: template_word.docx
csl: template_references_j_appl_ecol.csl
bibliography: bibliography.bib
---

@@

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Site-specific seed mixtures are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, reduce mowing efforts, and generate species-rich grasslands.
2.  A factorial field experiment addressed this topic on a dike at the Danube River in SE Germany in the 2018--2021 time period. On 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north or south exposition, and measured the recovery completeness by calculating the successional distance to reference sites, the persistence of sown species, and the Favourable Conservation Status (FCS) of target species in general.
3.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry grasslands.
4.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment. This initial negative effect remained over the entire observation period.
5.  Sand admixture had a slightly positive effect on target variables, while substrate depth, seed density and mixture type had no effects on species persistence or FCS.
6.  Synthesis and applications: Site-adapted seed mixtures can make restoration more efficient. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt, as well as re-sowing after unfavourable conditions. \clearpage \newpage

@@

Restoration ecology can increase the predictability of restoration approaches [@Suding.2011; @Mouquet.2015] by using rigorous, repeatable, and transparent experiments based on advanced theory, which will finally strengthen evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration method are key predictors for vegetation development after sowing [@Brudvig.2017b]. Hhabitat and biotic filtering are the main assembly factors which can be manipulated by the choice of seed--substrate combinations [@Torok.2017]. This means a close adaptation of the substrate to the niche of the target species or of the seed mixtures to the characteristics of the chosen substrate. Suitable substrates should reduce habitat filtering, while the seed mixture should minimize the competitive exclusion of unsuitable species. Modifying seed mixtures to match specific site conditions could be based on functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020], although this is not easy to implement [@merchant2022; @Bauer.2022]. This challenge is particularly interesting on artificial substrates that are used in urban areas [@Bauer.2022], quarries [@chenot-lescure2022], or on dikes [@Liebrand.1996].

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. Steep slopes with different exposition, contrasting substrate layers and dense swards for erosion protection characterize dikes [@Batori.2016; @berendse2015; @husicka2003]. Dikes can reconcile several ecosystem functions like dike security with biodiversity [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

The combinations of seed mixtures and substrates ('seed--substrate combinations') were tested on a dike at the Danube River in SE Germany (Figure \@ref(fig:map); 314 m a.s.l.; WGS84 (lat/lon), 48.83895, 12.88412). The climate of the region is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], as well as three minor floods which did not reach the plots [Appendix A1, @lfu2021waterlevel; @lfu2021discharge]. The substrates consisted of calcareous sand (0--4 mm) and agricultural soil obtained from a nearby dike construction site near the village of Steinkirchen. A big roller mixed both components and an excavator put the substrates in the prepared plots.

The target vegetation types were lowland mesic hay meadows and semi-dry calcareous grassland [EUNIS codes R22, R1A, @Chytry.2020, 'Arrhenatherion elatioris' and 'Cirsio-Brachypodion pinnati' according to the EuroVegChecklist, codes CM01A, DA01B, @mucina2016]. The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds (Co. Krimmer, Pulling). From these species pools, 20 species were selected for each plot in a stratified, randomised manner (Appendix A2). Each mixture contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Appendix A3). As expected, the hay-meadow mixtures had a higher specific leaf area (SLA), lower seed mass, and higher canopy height than the dry-grassland mixtures. The south-exposed plots were sown in mid-April 2018 and the north exposition 14 days later. Due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks due to unsatisfactory effects for the seedlings. The management started with a cut without hay removal and a mowing height of 20 cm in August 2018, followed by standard deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions.

We used 288 plots of the size 2.0 m × 3.0 m, vertically oriented, halfway up the dike slopes, distributed over the north and south exposition, and arranged in six blocks (= replicates). The experiment used a split-plot design combined with a randomized complete block design (Figure \@ref(fig:map)). The split plot was created by the two expositions of the dike, where all 24 treatment combinations were tested, i.e., sand admixtures (0, 25, and 50%), soil depths (15 vs. 30 cm), and the two seed mixture types with the normal seed density and an increased density recommended for slopes [4 vs. 8 g m^-2^, @Kiehl.2010; @KleberLerchbaumer.2017].

Under the substrate, a 5 cm thick drainage layer of gravel (0-16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced calcium carbonate, but did hardly change the pH which were within the weak alkaline range (Table \@ref(tab:substrate)). The pH corresponded to a well-established dike vegetation on the Rhine River [@husicka2003], phosphate and potassium were rather scarce for agricultural soils, but magnesium showed high concentrations [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and the Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A4. In total, 48 species of the species pool of hay meadows (87%) and 46 (79%) of dry grasslands established by 2021, which are rather good ratios [@hedberg2010]. The species established in 31 ± 22% (mean ± SD) of their sown plots. In total, 274 vascular plant species were detected (Appendix A5). The recovery completeness was described by the successional distance which quantifies the distance of a plot to the average reference site in the ordination [*d~jt,0~*, @Rydgren.2019]. Persistence was derived from the 'species losses' component of the temporal beta-diversity index (TBI; 1 − *B*~sor~) which was calculated by comparing the seed mixtures with the respective species composition of each year using Sørensen dissimilarity [@legendre2019]. The Favourable Conservation Status (FCS) is the ratio of characteristic and derived diversity, in our case species richness. Characteristic diversity consists of species which belong to a habitat-specific species pool and derived diversity consists of all other species. The habitat-specific species pool consisted of all sown species and other typical species of mesic and dry grasslands [Appendix A5, @Helm.2015].

Reference vegetation surveys were extracted from sPlotOpen [@Sabatini.2021] and our own surveys on the Danube dikes in the surroundings [@bauer2023]. We selected six calcareous grassland plots [EUNIS code R1A, @Chytry.2020] within SE Germany from sPlotOpen and 98 plots of our own survey which included also lowland hay meadows (R22) and as a negative reference ruderal, dry and anthropogenic vegetation (V38).

@@

A non-metric multidimensional scaling ordination (NMDS) with Sørensen dissimilarity was used to visualize variation in species composition in space and time (2D-stress 0.21). Seven species were excluded because they had an accumulated cover over all plots of less than 0.5%. Finally, 343 species were included in the ordination.

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists recording a certain plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used four chains, a thinning rate of two, 5,000 iterations for warm-up, and 10,000 in total. We used the Markov Chain Monte Carlo method (MCMC) with the no-U-turn Sampler (NUTS), which is based on the Hamiltonian Monte Carlo approach (HMC). For evaluating the computation, the convergence of the four chains was checked using trace plots and evaluating R-hat values, and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive checks were done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, the models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019].

@@

The ordination visualizes the species composition of seed mixtures and the development of the plots over time (Figure \@ref(fig:nmds)). The NMDS confirmed that the seed mixtures were variable, but distinctive for hay meadows and dry grasslands and ascertained the intended direction of the vegetation development. As one exception, hay meadows in south exposition did not develop towards their seed-mixture compositions.

@@

We could identify a statistically clear positive effect of the sand admixture on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed strongly positive after one year, but the recovery rate slowed down in the following years. Both expositions had similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong (Figures \@ref(fig:interactions)). Persistence of both seed mixture types was slightly positively affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

The vegetation of the dike experiment developed in the desired direction, but with a decelerating recovery rate. A significantly worse effect during establishment phase remained over the complete observation period in south compared to north exposition. All in all, it was hardly possible to identify any best seed--substrate combinations, although the hay meadows developed closer to their reference on the north exposition than the dry grasslands.

@@

The seed mixtures and their positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the NMDS suggests that the seed mixture represents a later successional stage than the references. The NMDS shows a slightly better adaptation of hay meadows to the north exposition than of dry grasslands (Figure \@ref(fig:interactions)A). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north-exposed dike slopes [@Batori.2020; @Oberdorfer.1993]. On the southern exposition, the plots of hay meadows developed rather to dry grassland references which indicates an inefficient restoration by using a non-adapted seed mixture.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral component need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to that study in France, sand admixture reduced nutrient loads and led to higher persistence of sown species while a 50% admixture did not further increase the effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the fourth year after restoration was rather small. The Favourable Conservation Status (FCS) was hardly affected by the sand admixture, which corresponds to the experiment in a quarry [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or FCS, similarly to earlier studies [@husicka2003; @baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010] or a substrate depth of lower 15 cm since most of the roots are in this soil layer [@vannoppen2016]. Seed density had also no clear effect on persistence and FCS which fits the results of @Kaulfuss.2022, who found that a certain amount of seeds is necessary for a successful establishment of target species, but higher densities do not further improve the outcome, but rather have a slightly negative effect.

The vegetation in south exposition had a different species composition, which confirms the findings of @Batori.2016 in Hungary. Here, the differences might be due to some unintended methodical reasons: the geotextile, which had been implemented on the southern slope, was removed after two weeks. This led to the drying of many seedlings which was amplified by the intense drought of 2018 and 2019 [@hari2020]. This drought could have decreased germination and the establishment of sown species, also generally speaking, because a moist spring with a mesic temperature enhances establishment [@Larson.2021; @orrock2023]. The interesting point here is the lasting negative effect on persistence and FCS on the southern slope which suggests a legacy effect by adverse weather conditions after sowing as it was observed by other studies [@Stuble.2017; @Groves.2020]. The problems during the establishment phase might have led to an alternative trajectory [@suding2004] and could probably only be met by re-sowing. The drying of seedlings might also have levelled the distinction of the seed mixture types on the south exposition.

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by sowing hay meadows in the north but not necessarily in the south exposition of dikes. Furthermore, the reduction of the nutrient load through sand admixture has a clearly positive effect, albeit the effect is small. The question remains if sand admixture is the most economical restoration measure to promote diversity on dikes. Increasing seed density on dike slopes does not appear to be necessary, and soil depths of 30 cm are not adverse compared to 15 cm thick substrates. There were no perfect seed--substrate combinations and thus we conclude that a variation of seed mixture types and different substrates along restoration sections would promote biodiversity better than a single uniform solution [@bauer2023; @Holl.2022]. The strong and lasting negative effects of the drought might require re-sowing. To close the recovery debt, the management adaptation might be promising since this is a crucial factor beside the restoration approach and the site characteristics [@Grman.2013; @tölgyesi2021]. For example, the start of sheep grazing on the experimental plots, which already exists in the surroundings, will modify the disturbance regime and improve dispersal. All in all, our results support the finding that restored dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. However, the recovery debt highlights the fact that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].

@@

```{r tab.id = "seedmix", ft.align = "left", tab.cap = "Each plot received an individual set of twenty species with some restrictions to the number of species per functional group. The total species pool for hay meadows was 55 and for dry grassland 58. In total, plots received 16 or 32 g of seeds. All individual seed mixtures are stored in Appendix A2."}
read_csv(here::here("data", "raw", "data_raw_table_1_seedmix.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  merge_at(part = "header", j = 2:3) %>%
  align(align = "center", part = "all", j = -1) %>%
  add_header_row(
    top = FALSE,
    values = c("", "Hay meadow", "Dry grassland", "", "", "")
    ) %>%
  add_header_row(top = FALSE, values = c("", "#", "#", "#", "wt%", "wt%")) %>%
  bold(part = "header", i = 1) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  autofit()
```

@@

Characteristics of the substrates used for the sowing experiment on river dikes. Soil samples of the three substrates were analyzed for the fraction \< 2 mm. The soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl~2~ solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in a CaCl~2~ extract. For calculating CaCO~3~, a sub-sample was annealed at 550 °C and the measured C amount multiplied with 8.33. To calculate total N and the C/N ratio, a sub-sample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand.

@@

Local setting and design of the multifactorial experiment on grassland sowing at dikes. The experiment was located on a dike at the Danube River in SE Germany. The 288 plots were allocated in six blocks (white squares on the upper photo) and on the north and south slope (central photo) [Both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted: sand admixture, substrate depth, seed density, and seed mixture types H and D (hay meadows, dry grasslands). The western half of a block had a shallow substrate depth and within this, half of the substrates had different sand admixtures. The photo on the bottom shows the northern slope of one block in 2021, four years after sowing (Photo: Markus Bauer).

@@

Effects of treatments on the development of sown grassland communities at a river dike. The posterior density distributions (grey) is calculated over all four surveyed years and both expositions. Shown are the medians, 66% and 95% credible intervals, which were derived from a Bayesian linear mixed-effects model (BLMM). (A) The persistence of sown species, (B) the Favourable Conservation Status (FCS) and (C) the recovery completeness compared to reference sites. The FCS is the ratio of target species to non-target species. Note that the zero lines indicate that both levels have equal values. This means, e.g., that hay meadows are closer to their reference than dry grasslands (C).

@@

The development sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with two different seed mixture types. Three indices are evaluated. (A) Recovery completeness (*d~jt~*~,0~): the zero lines indicate the mean position of the reference sites for each habitat type on the NMDS axis 1 (Figure \@ref(fig:nmds)). The grey area marks the standard deviation of the position of the reference sites. (B) Persistence of sown species: losses component of the temporal beta-diversity index (1 − *B*~sor~) (C) Favourable Conservation Status (FCS): the zero line indicates that target and non-target species are balanced. Positive values indicate that there are more target species. Shown are the medians and 95% credible intervals of the posterior distributions, which were derived from a Bayesian linear mixed-effects model (BLMM).

>>>>

---
title: "Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands"
author: Markus Bauer*, Jakob K. Huber, Johannes Kollmann
output:
  officedown::rdocx_document:
    reference_docx: template_word.docx
csl: template_references_j_appl_ecol.csl
bibliography: bibliography.bib
---

@@

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Site-specific seed mixtures are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, reduce mowing efforts, and generate species-rich grasslands.
2.  A factorial field experiment addressed this topic on a dike at the Danube River in SE Germany in the 2018--2021 time period. On 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north or south exposition, and measured the recovery completeness by calculating the successional distance to reference sites, the persistence of sown species, and the Favourable Conservation Status (FCS) of target species in general.
3.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry grasslands.
4.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment. This initial negative effect remained over the entire observation period.
5.  Sand admixture had a slightly positive effect on target variables, while substrate depth, seed density and mixture type had no effects on species persistence or FCS.
6.  Synthesis and applications: Site-adapted seed mixtures can make restoration more effective. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt, as well as re-sowing after unfavourable conditions. \clearpage \newpage

@@

Restoration ecology can increase the predictability of restoration approaches [@Mouquet.2015] by using rigorous, repeatable, and transparent experiments based on advanced theory, which will finally strengthen evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration method are key predictors for vegetation development after sowing [@Brudvig.2017b]. Hhabitat and biotic filtering are the main assembly factors which can be manipulated by the choice of seed--substrate combinations [@Torok.2017]. This means a close adaptation of the substrate to the niche of the target species or of the seed mixtures to the characteristics of the chosen substrate. Suitable substrates should reduce habitat filtering, while the seed mixture should minimise the competitive exclusion of unsuitable species. Modifying seed mixtures to match specific site conditions could be based on functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020], although this is not easy to implement [@merchant2022; @Bauer.2022]. This challenge is particularly interesting on artificial substrates that are used in urban areas [@Bauer.2022], quarries [@chenot-lescure2022], or on dikes [@Liebrand.1996].

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. Steep slopes with different exposition, contrasting substrate layers and dense swards for erosion protection characterise dikes [@Batori.2016; @berendse2015; @husicka2003]. Dikes can reconcile several ecosystem functions like dike security with biodiversity [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

The combinations of seed mixtures and substrates ('seed--substrate combinations') were tested on a dike at the Danube River in SE Germany (Figure \@ref(fig:map); 314 m a.s.l.; WGS84 (lat/lon), 48.83895, 12.88412). The climate of the region is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [Appendix A1, @hari2020], as well as three minor floods which did not reach the plots (Appendix A1). The substrates consisted of calcareous sand (0--4 mm) and agricultural soil obtained from a nearby dike construction site near the village of Steinkirchen. A big roller mixed both components and an excavator put the substrates in the prepared plots.

The target vegetation types were lowland mesic hay meadows and semi-dry calcareous grassland [EUNIS codes: R22, R1A, @Chytry.2020; Arrhenatherion elatioris and Cirsio-Brachypodion pinnati according to the EuroVegChecklist: CM01A, DA01B, @mucina2016]. The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds (Co. Krimmer, Pulling). From these species pools, 20 species were selected for each plot in a stratified, randomised manner (Appendix A2). Each mixture contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The hay-meadow mixtures had higher community-weighted means (CWM) for specific leaf area (SLA), lower for seed mass, and higher for canopy height than the dry-grassland mixtures (Appendix A3). The south-exposed plots were sown in mid-April 2018 and the north exposition 14 days later. Due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks due to unsatisfactory effects for the seedlings. The management started with a cut at 20 cm height without hay removal in August 2018, followed by standard deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions.

We used 288 plots of the size 2.0 m × 3.0 m, vertically oriented, halfway up the dike slopes, distributed over the north and south exposition, and arranged in six blocks (= replicates). The experiment used a split-plot design combined with a randomised complete block design (Figure \@ref(fig:map)). The split plot was created by the two expositions of the dike, where all 24 treatment combinations were tested, i.e., sand admixtures (0, 25, and 50%), soil depths (15 vs. 30 cm), and the two seed mixture types with the normal seed density and an increased density [4 vs. 8 g m^-2^, @Kiehl.2010; @KleberLerchbaumer.2017].

Under the substrate, a 5 cm thick drainage layer of gravel (0--16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced calcium carbonate, but did hardly change the pH which were within the weak alkaline range (Table \@ref(tab:substrate)). The pH corresponded to a well-established dike vegetation on the Rhine River [@husicka2003], phosphate and potassium were rather scarce for agricultural soils, but magnesium showed high concentrations [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and the Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A4. In total, 48 species of the species pool of hay meadows (87%) and 46 (79%) of dry grasslands established by 2021, which are rather good ratios [cf. @hedberg2010]. The species established in 31 ± 22% (mean ± SD) of their sown plots. In total, 274 vascular plant species were detected (Appendix A5). The recovery completeness was described by the successional distance which quantifies the distance of a plot to the average reference site in the ordination [*d~jt,0~*, @Rydgren.2019, Figure \@ref(fig:nmds)]. Persistence was derived from the 'species losses' component of the temporal beta-diversity index (TBI; 1 − *B*~sor~) which was calculated by comparing the seed mixtures with the respective species composition of each year using Sørensen dissimilarity [@legendre2019]. The Favourable Conservation Status (FCS) is the ratio of characteristic and derived diversity measured as species richness [@Helm.2015]. Characteristic diversity consists of species which belong to a habitat-specific species pool and derived diversity consists of all other species. The habitat-specific species pool consisted of all sown species and other typical species of mesic and dry grasslands (Appendix A5).

To compare the restoration outcomes with real references and not solely with seed mixtures, vegetation surveys were extracted from sPlotOpen [@Sabatini.2021] and our own surveys on the Danube dikes in the surroundings [@bauer2023]. We selected six dry grassland plots [EUNIS code R1A, @Chytry.2020] within SE Germany from sPlotOpen and 98 plots of our own survey which included also hay meadows (R22) and as a negative reference ruderal, dry and anthropogenic vegetation (V38).

@@

A non-metric multidimensional scaling ordination (NMDS) with Sørensen dissimilarity was used to visualise variation in species composition in space and time. Seven species were excluded because they had an accumulated cover over all plots of less than 0.5%. Finally, 343 species were included in the ordination.

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists recording a certain plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used four chains, a thinning rate of two, 5,000 iterations for warm-up, and 10,000 in total. We used the Markov Chain Monte Carlo method (MCMC) with the no-U-turn Sampler (NUTS). For evaluating the computation, the convergence of the four chains was checked using trace plots and evaluating *R*-hat values, and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive checks were done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, the models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019].

@@

The ordination visualises the species composition of seed mixtures and the development of the plots over time (Figure \@ref(fig:nmds)). The NMDS confirmed that the seed mixtures were variable, but distinctive for hay meadows and dry grasslands and ascertained the intended direction of the vegetation development. As one exception, hay meadows in south exposition did not develop towards their seed-mixture compositions.

@@

We could identify a statistically clear positive effect of the sand admixture on the persistence of sown species and on the recovery rate (Figure \@ref(fig:effects)). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed strongly positive after one year, but the recovery rate slowed down in the following years. Both expositions had similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong. Persistence of both seed mixture types was slightly positively affected by sand admixture on the north exposition (+ 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

@@

The seed mixtures and their positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the ordination suggests that the seed mixture represents a later successional stage than the references. The NMDS shows a slightly better adaptation of hay meadows to the north exposition than of dry grasslands (Figure \@ref(fig:interactions)A). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north-exposed dike slopes [@Batori.2020; @Oberdorfer.1993]. On the southern exposition, the plots of hay meadows developed rather to dry grassland references which indicates an ineffective restoration by using a non-adapted seed mixture.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral component need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to that study in France, sand admixture reduced nutrient loads and led to higher persistence of sown species while a 50% admixture did not further increase the effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the fourth year after restoration was rather small. The Favourable Conservation Status (FCS) was hardly affected by the sand admixture, which corresponds to the experiment in a quarry [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or FCS, similarly to earlier studies [@husicka2003; @baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010] or a substrate depth of lower 15 cm since most of the roots are in this soil layer on dikes [@vannoppen2016]. Seed density had also no clear effect on persistence and FCS which fits the results of @Kaulfuss.2022, who found that a certain amount of seeds is necessary for a successful establishment of target species, but higher densities do not further improve the outcome, but rather have a slightly negative effect.

The vegetation in south exposition had a different species composition, which confirms the findings of @Batori.2016 in Hungary. Here, the differences might be due to some unintended methodical reasons: the geotextile, which had been implemented on the southern slope, was removed after two weeks. This led to the drying of many seedlings which was amplified by the intense drought of 2018 and 2019 [@hari2020]. This drought could have also decreased germination and the establishment of sown species, because a moist spring with a mesic temperature enhances establishment [@Larson.2021; @orrock2023]. The lasting negative effect on persistence and FCS on the southern slope suggests a legacy or year effect by adverse weather conditions after sowing as it was observed by other studies [@Stuble.2017; @Groves.2020]. The problems during the establishment phase might have led to an alternative trajectory [@suding2004] and could probably only be met by re-sowing. The drying of seedlings might also have levelled the distinction of the seed mixture types on the south exposition.

@@

Our results suggest that adapted seed mixtures can increase restoration effectiveness by sowing hay meadows in the north but not necessarily in the south exposition of dikes. Furthermore, the reduction of the nutrient load through sand admixture has a clearly positive effect, albeit the effect is small. The question remains if sand admixture is the most economical restoration measure to promote diversity on dikes. Increasing seed density on dike slopes does not appear to be necessary, and soil depths of 30 cm are not adverse compared to 15 cm thick substrates. There were no perfect seed--substrate combinations and thus we conclude that a variation of seed mixture types and different substrates along restoration sections would promote biodiversity better than a single uniform solution [@bauer2023; @Holl.2022]. The strong and lasting negative effects of the drought might require re-sowing. To close the recovery debt, the management adaptation might be promising since this is a crucial factor beside the restoration approach and the site characteristics [@Grman.2013; @tölgyesi2021]. For example, the start of sheep grazing on the experimental plots, which already exists in the surroundings, will modify the disturbance regime and improve dispersal. All in all, our results support the finding that restored dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. However, the recovery debt highlights the fact that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020].

@@

Each plot received an individual set of twenty species with some restrictions to the number of species per functional group. The total species pool for hay meadows was 55 and for dry grassland 58. All individual seed mixtures are stored in Appendix A2.

```{r tab.id = "seedmix", ft.align = "left", tab.cap = ""}
read_csv(here::here("data", "raw", "data_raw_table_1_seedmix.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  merge_at(part = "header", j = 2:3) %>%
  align(align = "center", part = "all", j = -1) %>%
  add_header_row(
    top = FALSE,
    values = c("", "Hay meadow", "Dry grassland", "", "", "")
    ) %>%
  add_header_row(top = FALSE, values = c("", "#", "#", "#", "wt%", "wt%")) %>%
  bold(part = "header", i = 1) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  autofit()
```

@@

Characteristics of the substrates used for the sowing experiment on river dikes. Soil samples of the three substrates were analysed for the fraction \< 2 mm. The soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl~2~ solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in a CaCl~2~ extract. For calculating CaCO~3~, a sub-sample was annealed at 550 °C and the measured C amount multiplied with 8.33. To calculate total N and the C/N ratio, a sub-sample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand.

@@

Local setting and design of the multifactorial experiment on grassland sowing on dikes. The experiment was located on a dike at the Danube River in SE Germany. The 288 plots were allocated in six blocks (white squares on the upper photo) and on the north and south slope (central photo) [Both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted: sand admixture, substrate depth, seed density, and seed mixture types H and D (hay meadows, dry grasslands). The western half of a block had a shallow substrate depth and within this, half of the substrates had different sand admixtures. The photo on the bottom shows the northern slope of one block in 2021, four years after sowing (Photo: Markus Bauer).

@@

Effects of treatments on the development of sown grassland communities at a river dike. The posterior density distributions (grey) is calculated over all four surveyed years and both expositions. Shown are the medians, 66% and 95% credible intervals, which were derived from a Bayesian linear mixed-effects model (BLMM). Shown are (A) the recovery completeness compared to reference sites, (B) the persistence of sown species, and (C) the Favourable Conservation Status (FCS). The FCS is the ratio of target species to non-target species. Note that the zero lines indicate that both levels have equal values. This means, e.g., that hay meadows are closer to their reference than dry grasslands (A).

@@

The development of grassland communities at a river dike over four years after sowing. The plots had substrates with different sand admixtures and were sown with two different seed mixture types. Three indices are evaluated. (A) Recovery completeness (*d~jt~*~,0~): the zero lines indicate the mean position of the reference sites for each habitat type on the NMDS axis 1 (Figure \@ref(fig:nmds)). The grey area marks the standard deviation of the position of the reference sites (Figure \@ref(fig:nmds)). (B) Persistence of sown species: losses component of the temporal beta-diversity index (1 − *B*~sor~). (C) Favourable Conservation Status (FCS): the zero line indicates that target and non-target species are balanced. Positive values indicate that there are more target species. Shown are the medians and 95% credible intervals of the posterior distributions, which were derived from a Bayesian linear mixed-effects model (BLMM).
